Variant calling with transformers

### 2022-05-01 manuscript/manuscript.tex

\begin{abstract}
	Detection of germline variants in next-generation sequencing data is an essential element of modern genomics. Variant detection tools typically rely on a variety of statistical algorithms and heuristic techniques to identify variants. Here we describe a new approach that replaces the handcrafted statistical methods with a single, end-to-end deep learning model. Our model frames variant detection as a sequence-to-sequence modeling task, and employs a transformer-based architecture to translate alignment columns into two predicted haplotype sequences. Trained on whole genome sequences from Genome-in-a-Bottle, we demonstrate that this approach yields sensitivity and precision similar to other top-performing callers, and that 
\end{abstract}

@@

Variant discovery tools use a variety of statistical techniques to identify variants. Early callers, such as samtools / mpileup (Li et al. 2009) and the UnifiedGenotyper tool from the Genome Analysis ToolKit (GATK, De Pristo et al. 2011) examined each alignment column individually to determine genotype. Because these tools do not realign reads, they rely on the alignments produced by read mapping tools such as BWA (Li \& Durbin 2009) . Later tools, such as the HaplotypeCaller from the GATK and Strelka 2 (Kim et al. 2018) incorporated a re-alignment of reads in each region of interest, resulting in a significant improvement in the precision and specificity of variant calls. The HaplotypeCaller tool constructs de Bruijn graphs from k-mers present in the reads in a region, realigns reads to the graphs, and uses a pair Hidden Markov Model (HMM) to determine the most likely path through the graph.

@@

Training tensors were generated by selecting 150bp genomic regions in a biased manner described below. For each region, we fetch all overlapping reads and sort them by first aligned reference position. If more than 100 overlapping reads were found we randomly downsampled the set to not exceed the maximum size. For each base in the selected reads nine features were encoded; the first four were the one-hot encoded base call, followed by base quality, two flags indicating if base `consumed' reference base (i.e. was not an insertion) or consumed a sequence read base (was not a deletion), and additional flags indicating sequence read direction and clipping status.  No gap tokens or other special handling was performed for insertions or deletions. In addition to sequenced reads, the first row in each encoded region was the reference sequence. For this special row we inserted a base quality of 100 for every position and did not set any of the other flags. Resulting tensors had dimension $[g, r, 9]$, where $g$ represents genomic position, $r$ represents read, and 9 is the number of features.

Target / label sequences were produced by obtaining truth variants from the Genome-in-a-Bottle VCF files for each sample and inserting the variants into the reference sequence. Two sequences were generated for each region, one representing each haplotype. In situations where phasing of the variants was ambiguous, the reads in the sequenced sample were examined to determine phase status. Briefly, all possible genotypes (pairs of haplotype sequences) were generated and reads were aligned via Smith-Waterman to the possible haplotypes, and the highest scoring genotypes selected as the most likely phasing. This phasing procedure was only performed if multiple heterozygous variants were present in the region.

@@

Given an alignment file in BAM or CRAM format,

@@

We describe a new approach to the problem of detecting sequence variants in next-generation sequencing (NGS) data based on a single, end-to-end deep learning model. Our approach envisions variant detection as a sequence-to-sequence modeling problem, akin to language translation, and leverages the successful transformer architecture to accomplish the task. In contrast to other variant detection methods (...), our approach does not rely on handcrafted statistical techniques such as de Bruijn graphs, hidden markov models, or heuristic thresholding and instead learns to construct accurate haplotypes directly from aligned NGS reads.

>>>>

\begin{abstract}
	Detection of germline variants in next-generation sequencing data is an essential element of modern genomics. Variant detection tools typically rely on a variety of statistical algorithms and heuristic techniques to identify variants. Here we describe a new approach that replaces the handcrafted statistical methods with a single, end-to-end deep learning model. Our model frames variant detection as a sequence-to-sequence modeling task, and employs a transformer-based architecture to translate alignment columns into two predicted haplotype sequences. We demonstrate that this method  predicts variant genotypes accurately, phases nearby variants correctly, reconstructs complex variants, and incorporates local sequence context and variation into prediction confidence. Trained on whole genome sequences from Genome-in-a-Bottle, we demonstrate that this approach yields slightly lower sensitivity and precision than current top-performing callers,  
\end{abstract}

@@

Variant discovery tools use a variety of statistical techniques to identify variants. Early callers, such as samtools / mpileup (Li et al. 2009) and the UnifiedGenotyper tool from the Genome Analysis ToolKit (GATK, De Pristo et al. 2011) examined each alignment column individually to determine genotype. Because these tools do not realign reads, they rely on the alignments produced by read mapping tools such as BWA (Li \& Durbin 2009). Later tools, such as the HaplotypeCaller from the GATK and Strelka 2 (Kim et al. 2018) incorporated local re-alignment of reads in each region of interest, resulting in a significant improvement in the precision and specificity of variant calls. The HaplotypeCaller tool constructs de Bruijn graphs from k-mers present in the reads in a region, realigns reads to the graphs, and uses a pair Hidden Markov Model (HMM) to determine the most likely path through the graph.

@@

Training tensors were generated by selecting 150bp genomic regions in a biased manner described below. For each region, we fetch all overlapping reads and sort them by first aligned reference position. If more than 100 overlapping reads were found we randomly downsample the set to not exceed the maximum size. For each base in the selected reads nine features were encoded; the first four were the one-hot encoded base call, followed by base quality, two flags indicating if base `consumed' reference base (i.e. was not an insertion) or consumed a sequence read base (was not a deletion), and additional flags indicating sequence read direction and clipping status.  No gap tokens or other special handling was performed for insertions or deletions. In addition to sequenced reads, the first row in each encoded region was the reference sequence. For this special row we inserted a base quality of 100 for every position and did not set any of the other flags. Resulting tensors had dimension $[g, r, 9]$, where $g$ represents genomic position, $r$ represents read, and 9 is the number of features.

Target / label haplotype sequences were produced by obtaining truth variants from the Genome-in-a-Bottle VCF files for each sample and inserting the variants into the reference sequence. Two sequences were generated for each region, one representing each haplotype. In regions where phasing of the variants was ambiguous, the reads in the sequenced sample were examined to determine phase status. Briefly, all possible genotypes (pairs of haplotype sequences) were generated and reads were aligned via Smith-Waterman to the possible haplotypes, and the highest scoring genotypes selected as the most likely phasing. This phasing procedure procedure was only attempted for variants less than 100bp apart, otherwise the region was discarded.

@@

Given an alignment file in BAM or CRAM format, we first identify regions where a potential variant might exist, and then run forward passes of the model using a sliding window over the region. Any genomic position in which at least two reads contain a base that differs from the reference or an indel are flagged as potentially containing a variant. Positions closer than 100bp are merged into a single region.

Given a region containing suspected variants, we perform multiple overlapping forward passes of the model with step size $k$. On each forward pass the model produces two predicted haplotypes. Each haplotype is aligned via Smith-Waterman to the reference sequence, and any mismatching positions are converted to variant calls. We record the number of windows in which each variant was detected

@@

We describe a new approach to the problem of detecting sequence variants in next-generation sequencing (NGS) data based on a single, end-to-end deep learning model. Our approach envisions variant detection as a sequence-to-sequence modeling problem, akin to language translation, and leverages the successful transformer architecture to accomplish the task. In contrast to other variant detection methods (...), our approach does not rely on handcrafted statistical techniques such as de Bruijn graphs, hidden Markov models, or heuristic thresholding and instead learns to construct accurate haplotypes directly from aligned NGS reads.
